Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Remove all comments, add overview, adapt bibfile
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2920 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7ebd0a65b072806dfd1b1b818aa9217f1ba3d815

Many approaches to model circular data assume that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''.
%This is also the basis
%which early regression models for circular data are built on as they regress
%the parameter(s) of the von Mises distribution on covariates. 
One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution, is predicted by a linear combination of
covariates. 
\cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a 
restricted interval of length $2\pi$. 
%%%
%%%In order to restrict the circular response to an interval of length $2\,pi\
%%%\cite{Johnson+Wehrly:1978} refined this idea by linintroducing link functions transforming
%%%the linear predictor to the restricted interval of length $2\pi$. 
%\cite{Johnson+Wehrly:1978} took this idea one step further by
%plugging in a link function such that the values of the linear predictor are
%transformed to a restricted interval of length $2\pi$. 
%This generalized linear
%model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
%subsequently by \cite{Fisher:1993} introducing another specific transformation
%function which maps the values of the real line to the unit circle. 
This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing independent GLMs for either the 
location or scale parameter with appropriate link functions while keeping the other 
parameter constant. Additionally, they developed a combined heteroscedastic version 
where the location and scale parameter are alternately reestimated conditional on 
respective sets of covariates until convergence.
%Moreover, the presented methodology does not only allow for the mean direction to be modeled 
%depending on covariates while a constant estimate is obtained for the concentration parameter 
%but also for the vice versa version. For each of these two approaches a separate model is introduced, 
%however, they are also combined in a joint model (referred to as ``mixed model'').
%%Moreover, the presented methodology allows for both distribution parameters, the mean
%%direction and the concentration parameter of the von Mises distribution, to be
%%modeled depending on covariates, not only separately but also in a
%%combined model (also referred to ``mixed model''). 
%In the employed iterative optimization algorithm the two sets of regression coefficients used
%to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
%%To obtain proper initial values it is recommended to apply two separate models 
%%where either one of the parameters is fixed while the other is predicted employing covariates 
%%in a GLM-type model.
%While the combined model but also the two separate models are built on well elaborated theory 
While all of these models are built on well elaborated theory 
the employed methodology has shown to be very challenging in its application, mainly due to the 
complexity encountered in optimizing the corresponding log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for these circular GLMs to converge 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
have presented a Bayesian alternative of a homoscedastic GLM for circular data.
However, apart from potential difficulties in the optimization procedure of circular GLMs, 
additive effects and their interpretation are often not intuitive for circular responses
due to the highly non-linear link function and the not-straightforward representation of smooth 
transitions on the unit circle. For example, the same rotation can be obtained in either 
positive or negative direction on the circle leading to an ambiguous interpretation.

%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
%%%are built on additive regression models.
%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
%%%however, in the case of modeling circular data 
%%%additive effects conditional on linear covariates are not intuitive
%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
%%%can be obtained in either positive or negative direction on the circle.
%a possibly highly non-linear link function
%can induce a more complex behavior of the modeled parameters as the location
%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
%direction of the shift.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link functions must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of statistical inference.
In addition, the employed tree structure allows to capture non-additive effects while forests 
enable the modeling of smooth changes.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

%Moreover, non-additive effects are captured by the tree structure while forests allow for 
%modeling smooth changes as well. Thus, a high level of flexibility is provided while a full 
%distributional model is specified in each terminal node offering a wide range of inference.

%The resulting circular regression trees and forests do not require a pre-specified link function 
%or any other model assumptions as covariates and their possible interactions are
%selected automatically.

%Moreover, as covariates are employed for splitting only and not
%considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach does not only
allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
%By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.

%Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
%of tree-based circular models by incorporating the framework of distributional modeling and in that way
%considering splits in all distribution parameters and providing a fully specified distribution fitted 
%in each terminal node.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%An alternative modeling approach that can deal with non-additive effects is
%provided by tree-based methods. \cite{Lund:2002} introduced the concept of
%circular regression trees which was further extended by
%\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
%methods but also including the probabilistic concept of regressing distribution
%parameters on covariates, we propose a very flexible approach by applying the
%von Mises distribution within the methodology of distributional trees and
%distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
%data. The resulting circular distributional trees and forests do not require
%any model assumptions as covariates and their possible interactions are
%selected automatically. Non-additive effects are captured by the tree structure
%while circular forests allow for modeling smooth changes as well. Thus, a high
%level of flexibility is provided while a full distributional model is specified
%offering a wide range of inference.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and \citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

In most generalized linear or additive approaches to circular regression modeling 
the location parameter $\mu$ is regressed on covariates $\mathbf{z}$ in the form of 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$; whereby the 
additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
As an appropriate link function typically $g(x) = 2 \cdot arctan(x)$ is employed, 
as suggested by \citet{Fisher+Lee:1992}.
While most of the proposed models estimate a constant concentration parameter, hence, assume
homoscedasticity, \citet{Fisher+Lee:1992} also developed a heteroscedastic version by combining two  
individual GLMs, each for one of the parameters $\mu$ and $\kappa$. 
This provides a first approach to a fully probabilistic regression model for circular 
data, however, the parameters are not regressed simultaneously on covariates as known from the 
more general framework provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. 
Nevertheless, all circular additive models incorporate the previously discussed difficulties induced by
the characteristics of the log-likelihood function and the strongly non-linear link function.
%For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
%on an interval of length $2\pi$ by setting 
%$g(x) = 2\pi \cdot F(x)$ 
%for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
%$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
%Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
%as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
%covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
%constant value. Additionally to these two model variations where only one of the two distribution 
%parameters depends on covariates a combination of them is introduced. In this joint model the 
%two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
%in alternation within an iterative optimization algorithm. 
%Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
%data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
%more general case by generalized additive models for location, scale, and shape 
%\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
%However, in practice the incorporated optimization procedure has shown to be very complex, already
%for only one parameter being regressed on covariates and consequently even more so in the joint model.
%This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
%distribution with the plugged-in link function. In particular, the function is not globally concave and
%possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
%Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
%is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
%covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
%compression of the estimated density function.
%\newline
Referring to additive models in general, it has to be considered that a proper model specification 
can be very challenging, particularly for a high number of covariates and no information on possible 
interactions. Moreover, the additive structure might impose a smooth effect even if the true underlying 
effect is an abrupt shift.

@@

%A very flexible framework should allow for detecting a wide range of 
%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
%to a circular representation should be avoided as well as any dependence on initial values.
%\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
%variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
%and where to split the data an angular distance measure is evaluated between the mean direction 
%and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
%Hence, only the location parameter is considered for splitting such that effects in other parameters 
%might be missed. Therefore, we introduce a distributional approach to circular regression trees 
%where a fully specified distributional model is fitted in each node of the tree which enables the 
%detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
%corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
%quantities.
%Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
%within each node is fitted without employing covariates.
%A common way of including covariates is to regress the model parameters on
%covariates. In the case of a probabilistic model for a circular response this
%corresponds to expressing the location parameter $\mu$ and the concentration
%parameter $\kappa$ of the von Mises distribution by means of the available
%covariates, e.g., in the form of a linear combination, possibly with an
%appropriate link function plugged in as in GLMs. This kind of approach is
%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
%\cite{Fisher:1993}. Apart from the methodology developed specifically for
%circular data more general approaches to regressing model parameters on
%covariates are commonly used such as for example generalized additive models
%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
%where each distribution parameter is estimated by its own generalized additive
%model (GAM). However, difficulties might arise in case of non-additive changes
%and a proper model specification can be very challenging, particularly for a
%high number of covariates and no information on possible interactions. As an
%alternative regression approach which can deal with these problems, we
%introduce a circular probabilistic tree-structured model in this section.

@@

\subsection{Data}
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airport for Innsbruck and Vienna on an
hourly temporal resolution. As predictor variables, we use temporal information
and $1$-hourly resolved $10$\,min mean observations of various meteorological quantities such as 
wind direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within an area of approximately 30\,km$^2$ are used. A topographical overview of
the airports and their surrounding areas with the station sites employed in
this study is provided in Figure~XXX. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

@@

    For simplicity, rather to fit one single climatological model, separate models are estimated 
    rolling over all observation time points, 
    using the same daytime of $31$~days centered around the day of interest over all available years in the training data.
    %For simplicity, rather to fit one single climatological model, we perform the
    %model fitting separately rolling over all time points. For each time point, the
    %training data set consists of the same daytime as the time of interest within
    %31~days centered around the same calender day over all years except the year of
    %interest. 
    To gain full probabilistic climatological forecasts, we employ the
    steps described in Section~\ref{sec:prob_circ} and fit the distribution
    parameters of the von Mises distribution via maximum likelihood estimation. The
    construction of a climatological reference model was similarly pursued by
    \citet{Vogel+Knippertz+Fink:2018} and is discussed in a comprehensive summary on
    different time-adaptive training schemes in \citet{Lang+Lerch+Mayr:2019}.

@@

    To gain a full probabilistic persistency model, we proceed similarly as for the
    climatological model by using maximum likelihood estimation and fitting the

@@

    We fit one model for every hour throughout 
    the validation data set employing the
    previous six lagged response values as training data. In order to allow for a
    stronger influence of observations closer to the time of interest, exponential
    smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
    prediction an equal influence rate of 50 percent is assigned both to the current 
    observation and to the previous five observations together. Observations with longer 
    time lags have exponential weights below $0.01$ and are therefore omitted from the
    training data.

@@

    Traditional approaches to forecast circular response variables are often based
    on circular GLM-typed models \citep{Fisher:1993}. As discussed in

@@

    learning sample. To allow for seasonally varying error characteristics in both bias and slope
    coefficients, and to allow for seasonal heteroscedasticity captured by the concentration
    parameter, we use the same time-adaptive trainings approach as for the
    climatological model; hence, separate models are estimated over all observation time points,
    using the same daytime of $31$~days centered around the day of interest over
    all available years in the training data \citep{Lang+Lerch+Mayr:2019}.

@@

    Section~\ref{sec:circforest}, the circular forest used in this study is constructed 
    based on $100$~individual trees employing the R-package \pkg{circtree}. Each of these 
    trees is again built by the CTree algorithm on a subsample containing a fraction of 
    $30$~percent of the original learning data. As it can be expected that the lagged
    response variable is one of the most informative predictor variables for wind
    direction forecasting, all covariates are included for building each tree which
    ensures that the lagged response variable is always considered for splitting.
    This bagging approach can be applied in \pkg{circtree} by setting the argument
    \code{mtry} to the total number of covariates. Since a high number of possible
    split points leads to high computational costs, the covariates are binned in a
    maximum of $50$ classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}).
    Contrary to a single-tree model, forests usually consist of very large trees as
    they are not prone to overfitting the data due to the stabilization obtained by
    combining the individual trees. Therefore, we use the following control
    arguments to build rather large trees: The minimal number of observations to
    perform a split is set to 20 (argument \texttt{minsplit}), the minimal number
    of observations in each terminal node is set to 7 (argument
    \texttt{minbucket}), and the significance level for variable selection is kept
    at its maximum value of $1$ (argument \texttt{alpha}).

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and forests 
are established by coupling model-based recursive partitioning with the von Mises distribution. 
By separating the data into more homogeneous subgroups, possible difficulties in circular regression
are bypassed as covariates are solely considered for splitting and 
group-specific models are fitted as intercept-only models.
%distributional models are fitted as intercept-only models in each node. 
In addition, by specifying the von Mises distribution for each node and
allowing for splits in both distribution parameters $\mu$ and $\kappa$, 
fully probabilistic forecasts are provided.

@@

%The comparison of the novel circular trees with simple climatological and persistency models, depicted
%for probabilistic wind direction forecasts at two airports, shows that even in case of no obvious 
%skillful covariates the trees have an high predictive performance.

%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{itemize}
%  \item Add summary/conclusion of case study;
%  \item Add summary/conclusion of circular regression trees and forests.
%  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
%\end{itemize}

>>>>

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
the location parameter of the von Mises distribution, is predicted by a linear
combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM)-type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
version where the location and scale parameter are alternately reestimated
conditional on respective sets of covariates until convergence. While all of
these models are built on well elaborated theory the employed methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly informative starting values are
crucial for these circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} have presented a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, additive effects
and their interpretation are often not intuitive for circular responses due to
the highly non-linear link function and the not-straightforward representation
of smooth transitions on the unit circle. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and forests
\citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
trees and forests avoid the discussed difficulties of circular GLMs by
separating the data into more homogeneous subgroups and fitting a simple
intercept-only model for each resulting subset. Therefore, no covariates and,
thus, no link functions must be employed in the estimation of the
group-specific distribution parameters which also simplifies the optimization
process. By incorporating the framework of distributional modeling, we allow
for detecting splits in all distribution parameters and provide a fully
specified circular response distribution for each terminal node offering a wide
range of statistical inference. In addition, the employed tree structure
allows to capture non-additive effects while forests enable the modeling of
smooth changes. Furthermore, covariates and their possible interactions do not
need to be specified in advance as they are selected automatically.

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} has already
introduced the concept of circular regression trees where binary splits are
made based on an angular distance measure employed to evaluate node
homogeneity. While this approach does not only allow for linear, but also for
circular covariates, the tree solely considers effects in the mean direction
for splitting, i.e., only the location parameter $\mu$ is regressed on
covariates. Therefore, the introduced distributional approach to circular
regression trees extends the concept of previous circular trees by considering
splits in all distribution parameters and providing a full probabilistic model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study.

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

In most generalized linear or additive approaches to circular regression
modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
in the form of 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$;
whereby the additive predictor is transformed to an interval of length $2\pi$
by a link function $g$. As an appropriate link function typically $g(x) = 2
\cdot arctan(x)$ is employed, as suggested by \citet{Fisher+Lee:1992}. While
most of the proposed models estimate a constant concentration parameter, hence,
assume homoscedasticity, \citet{Fisher+Lee:1992} also developed a
heteroscedastic version by combining two individual GLMs, each for one of the
parameters $\mu$ and $\kappa$. This provides a first approach to a fully
probabilistic regression model for circular data, however, the parameters are
not regressed simultaneously on covariates as known from the more general
framework provided by generalized additive models for location, scale, and
shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular
additive models incorporate the previously discussed difficulties induced by
the characteristics of the log-likelihood function and the strongly non-linear
link function. Referring to additive models in general, it has to be
considered that a proper model specification can be very challenging,
particularly for a high number of covariates and no information on possible
interactions. Moreover, the additive structure might impose a smooth effect
even if the true underlying effect is an abrupt shift.

@@

@@

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for the airports Innsbruck and Vienna. 
    For Innsbruck, $4$~stations at the airport and $6$~along the
    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
    and $13$~within an area of approximately 30\,km$^2$ are used.
    Elevation data are obtained from the TandDEM-X digital elevation model with a 
    horizontal resolution of $90$~m \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airport for Innsbruck and Vienna on
an hourly temporal resolution. As predictor variables, we use temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities such as wind direction, wind speed, temperature, air
pressure and humidity, all lagged by one or three hours according to the
respective forecasting step. The meteorological variables are measured either
directly at the airports or within their vicinities. For Innsbruck,
measurements at the airport and along the intersecting valleys are used,
whereas, for Vienna, measurements at the airport and within an area of
approximately 30\,km$^2$ are used. A topographical overview of the airports and
their surrounding areas with the station sites employed in this study is
provided in Figure~\ref{fig:overview}. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

@@

    For simplicity, rather to fit one single climatological model, separate
    models are estimated rolling over all observation time points, using the same
    daytime of $31$~days centered around the day of interest over all available
    years in the training data.  To gain full probabilistic climatological
    forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
    the distribution parameters of the von Mises distribution via maximum
    likelihood estimation. The construction of a climatological reference model was
    similarly pursued by \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
    comprehensive summary on different time-adaptive training schemes in
    \citet{Lang+Lerch+Mayr:2019}.

@@

    To gain a full probabilistic persistency model, we proceed similarly as for
    the climatological model by using maximum likelihood estimation and fitting the

@@

    We fit one model for every hour throughout the validation data set employing
    the previous six lagged response values as training data. In order to allow for
    a stronger influence of observations closer to the time of interest,
    exponential smoothing is employed with a smoothing factor of $0.5$;
    accordingly, for every prediction an equal influence rate of 50 percent is
    assigned both to the current observation and to the previous five observations
    together. Observations with longer time lags have exponential weights below
    $0.01$ and are therefore omitted from the training data.

@@

    Traditional approaches to forecast circular response variables are often
    based on circular GLM-typed models \citep{Fisher:1993}. As discussed in

@@

    learning sample. To allow for seasonally varying error characteristics in both
    bias and slope coefficients, and to allow for seasonal heteroscedasticity
    captured by the concentration parameter, we use the same time-adaptive
    trainings approach as for the climatological model; hence, separate models are
    estimated over all observation time points, using the same daytime of $31$~days
    centered around the day of interest over all available years in the training
    data \citep{Lang+Lerch+Mayr:2019}.

@@

    Section~\ref{sec:circforest}, the circular forest used in this study is
    constructed based on $100$~individual trees employing the R-package
    \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a
    subsample containing a fraction of $30$~percent of the original learning data.
    As it can be expected that the lagged response variable is one of the most
    informative predictor variables for wind direction forecasting, all covariates
    are included for building each tree which ensures that the lagged response
    variable is always considered for splitting. This bagging approach can be
    applied in \pkg{circtree} by setting the argument \code{mtry} to the total
    number of covariates. Since a high number of possible split points leads to
    high computational costs, the covariates are binned in a maximum of $50$
    classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}). Contrary to a
    single-tree model, forests usually consist of very large trees as they are not
    prone to overfitting the data due to the stabilization obtained by combining
    the individual trees. Therefore, we use the following control arguments to
    build rather large trees: The minimal number of observations to perform a split
    is set to 20 (argument \texttt{minsplit}), the minimal number of observations
    in each terminal node is set to 7 (argument \texttt{minbucket}), and the
    significance level for variable selection is kept at its maximum value of $1$
    (argument \texttt{alpha}).

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and
forests are established by coupling model-based recursive partitioning with the
von Mises distribution. By separating the data into more homogeneous
subgroups, possible difficulties in circular regression are bypassed as
covariates are solely considered for splitting and group-specific models are
fitted as intercept-only models. In addition, by specifying the von Mises
distribution for each node and allowing for splits in both distribution
parameters $\mu$ and $\kappa$, fully probabilistic forecasts are provided.

@@
